---
title: "Global GDP per Capita Analysis (2024)"
author: "Girish R"
format: html
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(scales)
```

# GDP per Capita Analysis

## Introduction

This report analyzes global GDP per capita (current US\$) in 2024, identifying the wealthiest and poorest nations. Data sourced from the World Bank (hypothetical example).

## Data Loading & Cleaning

```{r load-data}
gdp <- read_csv("gdp_2024.csv") %>%
  select(country, obs_value) %>%
  rename(gdp_per_capita = obs_value) %>%
  arrange(desc(gdp_per_capita))
```

## Top 10 Countries

```{r top-10}
top_10 <- gdp %>% slice_max(gdp_per_capita, n = 10)

ggplot(top_10, aes(x = reorder(country, gdp_per_capita), y = gdp_per_capita)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(title = "Top 10 Countries by GDP per Capita (2024)",
       x = "", y = "USD") +
  scale_y_continuous(labels = comma) +
  theme_minimal()
```

**Narrative:**\
Monaco leads with \$240,862 per capita, followed by Liechtenstein and Luxembourg. These microstates benefit from favorable tax policies and financial services. The United States ranks outside the top 10 but remains a major economy at \$81,695.

## Bottom 10 Countries

```{r bottom-10}
bottom_10 <- gdp %>% slice_min(gdp_per_capita, n = 10)

ggplot(bottom_10, aes(x = reorder(country, -gdp_per_capita), y = gdp_per_capita)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  labs(title = "Bottom 10 Countries by GDP per Capita (2024)",
       x = "", y = "USD") +
  scale_y_continuous(labels = comma) +
  theme_minimal()
```

**Narrative:**\
Burundi (\$199.58) and the Central African Republic (\$445.03) face significant challenges including political instability and limited infrastructure. All bottom 10 countries are in Africa, highlighting regional disparities.

## Global Distribution

```{r boxplot}
ggplot(gdp, aes(y = gdp_per_capita)) +
  geom_boxplot(fill = "skyblue") +
  scale_y_log10(labels = comma) +
  labs(title = "Global GDP per Capita Distribution (Log Scale)",
       y = "USD (log scale)") +
  theme_minimal()
```

**Narrative:**\
The logarithmic boxplot reveals extreme inequality – the top 25% of countries have GDP per capita values exceeding \$30,000, while the bottom 25% fall below \$2,000.

## Conclusion

Global wealth distribution remains highly unequal. Microstates dominate the top ranks, while sub-Saharan African nations are disproportionately represented in the lowest ranks.
